---
title: "Beginning Exploratory Data Analysis"
author: "Will & Sam"
date: "6/2/2017"
output: html_document
---

We have data for planting density and corresponding yield. We may incorporate other dependent variables as necessary. 

```{r}
library(ggplot2)

ggplot(data = , aes(x = planting density, y = yield, group = ))+
  geom_smooth(se = FALSE, alpha=0.25, size = 0.5)+
  geom_point(alpha = 0.25, size =0.5, position ='jitter')+
  geom_boxplot(se = FALSE)+
  
#geom_smooth
#geom_boxplot
#geom_violin
#geom_histogram
#ggthemes:theme_tufte()
  



```

```{r}
#Some statistics. Will need to be updated as actual data becomes available. 
traits_variables_season_2 %>%
  group_by(date) %>%
  summarize(mean = mean(mean), n = n())

traits_variables_season_2 %>%
  group_by(date, cultivar_id) %>%
  summarize(mean = mean(mean), n = n())

canopy_heights <- traits_variables_season_2 %>% 
  group_by(date, cultivar_id) %>% 
  summarize(mean = mean(mean), n =n()) 



```

Would also be interesting to compare data across and between different species. 

##Update 6/1/2017

We performed some beginning data analysis on '/data/yieldVersusDensity.rds' We observed that there are very few distinct planting densities for each different plant species. In fact, removing all NAs there are only 6 in the above table, and if we look at the larger yields database, there are 37 unique plant species, but only 33 unique planting densities. We are concerned that this is not enough data to obtain meaningful (since our question essentially reduces to "does species play a role in determining yield?", which seems almost tautological). 